MCFA on Galah

### 2018-11-04 article/ms.tex
boom

\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

@@

@@

\noindent{}where $\factorloads$ is a $p \times d$ matrix of factor
loads that is common to all data points, and the factor scores for
the $i$th data point
\begin{equation}
	\factorscores_i \sim \mathcal{N}(\vec\xi_k, \vec\Omega_k)
\end{equation}
\noindent{}are drawn from the $k$th multivariate  normal distribution.
The factor scores for all data points $\factorscores$ is then a 
\todo{$n \times p$} matrix.
 We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \specificvariance\right)$
is independent of the latent space, and $\specificvariance$ is a
diagonal matrix of $d$ entries.

@@

We generated a data set with $\todo{N = 100000}$ data points, each with
$D = 15$ dimensions, and assumed that those data are generated by
latent space of $J = 5$ ($J \times D$) factor loads, and there are $K = 20$
clusters in the latent space. The relative weights $\vec\weight$
are drawn from a multinomial distribution and the means of the clusters
in factor scores $\scoremeans$ are drawn from a standard normal
distribution. The covariance matrices in factor scores $\scorecovs$ are assumed to be \todo{low rank}, where the diagonal entries are drawn from a gamma distribution $\scorecovs_{k,i,i} \sim \Gamma(1)$. The variance in 
each dimension $\specificvariance$ are also drawn $\specificvariance \sim \Gamma(1)$.
The $i$th data point (which belongs to the $k$th cluster) is then
generated by drawing $\factorscores_{i} \sim \mathcal{N}(\scoremeans_k,\scorecovs_k)$, projecting by the factor loads, and adding variance.

Latent factor models are by definition indeterminate in that there are
multiple solutions of the parameter vector $\vec\Psi$ which provide
an identical data-generating process. This can be trivially shown by
the fact that the factors loads and scores can be rotated with some rotation
matrix $\textbf{R}$ and generate the exact same data. As a consequence,
we cannot strictly evaluate our model fitting procedure by comparing
how closely the inferred factor loads (or scores) match the true 
variables. This is further complicated by \todo{orientation and order}
of the latent factors: even if the inferred factor loads were very close
to the true factor loads, they could be ordered differently, or have
their signs reversed. Given these issues, we evaluate the search as
being successful if the log probability given the estimated model
parameters is within convergence tolerance (or better) than the
log probability given the data and the true parameters.

We fit this generated data using the expectation-maximization algorithm
as described in the previous Section. After \todo{X} iterations
(and a serial processing time of \todo{X} seconds) we
reached our convergence threshold: the log probability increased by
less than \todo{X} with each iteration. For this experiment we specified
the number of latent factors $J$ and the number of clustering components
$K$ to include, but below we repeat this experiment where the number
of latent factors and clustering components is not known.

We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
In this figure we have over-plotted \todo{100000} samples generated by the
estimated factor loads $\factorloads_{est}$ and the clustering in factor 
scores $\scoremeans_{est}$ and $\scorecovs_{est}$ (weighted by $\vec\weight_{est}$)
-- without adding variance $\specificvariance$ -- in order to illustrate that the
estimated model parameters can realistically generate the data.
This is despite the estimated factor loads being slightly different
than the true factor loads (Figure~\ref{fig:toy-model-factor-loads}),
a consequence of the uniqueness of latent factor models. The estimated
latent space of factor scores is shown in Figure~\ref{fig:toy-model-factor-scores}
where the estimated clustering of latent scores is shown.
% Despite not matching perfectly, it generates the data? property of uniqueness?

@@

@@

\acknowledgements 
\todo{
\begin{itemize}
	\item ARC DP
	\item GALAH acknowledgement
	\item Gaia acknowledgement
\end{itemize}
}

\software{
	\package{Astropy} \citep{astropy:v1,astropy:v2},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy},
    \package{Stan} \citep{stan},
    \package{CosmoHub} \citep{cosmohub},
    \package{TensorFlow} \citep{tensorflow}
    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
}

>>>>

\newcommand{\Gaia}{\project{Gaia}}
\newcommand{\gaia}{\project{gaia}}
\newcommand{\Galah}{\project{Galah}}
\newcommand{\GALAH}{\project{GALAH}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

@@

% TODO: Is D a diagonal vector, or a matrix?! Make it consistent.

@@

\noindent{}where $\factorloads$ is a $p \times d$ matrix of factor
loads that is common to all data points, and the factor scores for
the $i$th data point
\begin{equation}
	\factorscores_i \sim \mathcal{N}(\vec\xi_k, \vec\Omega_k)
\end{equation}
\noindent{}are drawn from the $k$th multivariate  normal distribution.
The factor scores for all data points $\factorscores$ is then a 
\todo{$n \times p$} matrix.
 We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \eye\specificvariance\right)$
is independent of the latent space, and $\specificvariance$ is a
diagonal matrix of $d$ entries.

@@

We generated a data set with ${N = 100000}$ data points, each with
$D = 15$ dimensions, and assumed that those data are generated by
latent space of $J = 5$ ($J \times D$) factor loads, and there are $K = 20$
clusters in the latent space. The relative weights $\vec\weight$
are drawn from a multinomial distribution and the means of the clusters
in factor scores $\scoremeans$ are drawn from a standard normal
distribution. The covariance matrices in factor scores $\scorecovs$ are assumed to be diagonal matrices, where the non-zero entries are drawn from a gamma distribution $\scorecovs_{k,i,i} \sim \vec\Gamma\left(1\right)$. The variance in 
each dimension $\specificvariance$ are also drawn $\specificvariance \sim \vec\Gamma\left(1\right)$.
The $i$th data point (which belongs to the $k$th cluster) is then
generated by drawing $\factorscores_{i} \sim \mathcal{N}(\scoremeans_k,\scorecovs_k)$, projecting by the factor loads, and adding variance.

We fit this generated data using the expectation-maximization algorithm
as described in the previous Section. After 58 iterations
(and a serial processing time of 40 seconds) we
reached our convergence threshold: the log probability increased by
less than $10^{-5}$ with each iteration. For this experiment we specified
the number of latent factors $J$ and the number of clustering components
$K$ to include, but below we repeat this experiment where the number
of latent factors and clustering components is not known.

We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
In this figure we have over-plotted 100000 samples generated by the
estimated factor loads $\factorloads_\textrm{est}$ and the clustering in factor 
scores $\scoremeans_\textrm{est}$ and $\scorecovs_\textrm{est}$ (weighted by $\vec\weight_\textrm{est}$)
-- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
-- in order to illustrate that the
estimated model parameters can realistically generate the data.
The estimated latent space of factor scores is shown in 
Figure~\ref{fig:toy-model-latent-space} where the colours indicate the
most likely component that the data point is associated with.

Latent factor models are by definition indeterminate in that there are
multiple solutions of the parameter vector $\vec\Psi$ which provide
an identical data-generating process. This can be trivially shown by
the fact that the factors loads and scores can be rotated with some rotation
matrix $\textbf{R}$ and generate the exact same data. As a consequence,
we cannot strictly evaluate our model fitting procedure by comparing
how closely the inferred factor loads (or scores) match the true 
variables. This is further complicated by orientation and order
of the latent factors: even if the inferred factor loads were very close
to the true factor loads, they could be ordered differently, or have
their signs reversed. Given these issues, we evaluate the search as
being successful if the log probability given the estimated model
parameters is within convergence tolerance (or better) than the
log probability given the data and the true parameters.

% Despite not matching perfectly, it generates the data? property of uniqueness?
% Factor scores?
% Some metric about clustering performance against true values?

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/toy-model-data.png}
    \caption{\todo{Caption}}
    \label{fig:toy-model-data}
\end{figure*}

\begin{figure*}
	\includegraphics[width=1.0\textwidth]{experiments/toy-model-latent-space.png}
    \caption{\todo{Caption}}
    \label{fig:toy-model-latent-space}
\end{figure*}

% Search strategy for toy model.

% Figure: toy-model-factor-loads?
% Figure: toy-model-specific-variances?

@@

The \Galah\ survey is a stellar spectroscopic survey of up to
$10^6$ stars. Up to 30 chemical abundances are expected to be measured
and reported in the final \Galah\ data release.
\todo{more about galah}

We use the second \Galah\ data release \citep{Buder:2018a} which
includes up to \todo{X} chemical abundances measured for \todo{Y}
stars. We selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances are untrustworthy. For this subset
we examined the measured abundances of \todo{X, etc} and
counted the number of stars with \texttt{flag\_x\_fe = 0}
for each abundance. This flag indicates that there is no known
reason to suspect that the measured chemical abundance is
unreliable.

We sorted the elemental abundances by the number of stars
with reliable measurements, from highest to lowest. Starting
with Fe, we iterated over each element and counted the number
of stars with reliable measurements of that element and all
other elements earlier in the list. We illustrate this process
in Figure~\ref{fig:galah-abundance-flags}, where we show that
if we include the top 15 elemental abundances, we would have
\todo{X} stars with reliable measurements in all fifteen elements.
This is relevant to the approach we use here, as the method
described here requires no missing data.

% Where do we go down to?
% What elements do we add /remove.
\todo{Where do we go down to? What elements do we exclude?}
This leaves us with a sample of \todo{X} stars each with
reliable stellar parameters ($\teff$, $\logg$) and 
\todo{Y} (reliable) measured abundances: \todo{Mn, Fe}.
These elements are produced from multiple nucleosynthetic
pathways, including \todo{X}. 
It is this subsample that we will use for experiments
in chemical tagging.

\subsubsection{Search strategy} \label{sec:experiment-galah-search}

% Unknown number of latent factors and unknown number of latent scores

@@

\acknowledgements
A.~R.~C. is supported in part by Australian Research Council
Discovery Project DP160100637.
% Gaia acknowledgement?
% GALAH acknowledgement?

\software{
	\package{Astropy} \citep{astropy:v1,astropy:v2},
    \package{IPython} \citep{ipython},
    \package{matplotlib} \citep{mpl},
    \package{numpy} \citep{numpy},
    \package{scipy} \citep{scipy},
    \package{Stan} \citep{stan},
    \package{Jupyter Notebooks} \citep{jupyter-notebooks}
}
